Density-dependent selection and the limits of relative fitness

### 2018-07-12 manuscript.tex
l

In the special case that all types are competitively equivalent (identical $c_i$), Eq.~\eqref{eq:master} takes a simpler form,
\begin{equation}
\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U=\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N), \label{eq:masterequalc}
\end{equation}
where we have used the fact that $L=\overline{b}N/T$ to make the dependence on $b$ and $N$ explicit ($\overline{b}$ is the population mean $b$). This formula can also be deduced directly from Eq.~\eqref{eq:growthsumuncoupled}, and is exact: $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of such territories, and type $i$ is expected to receive a fraction $l_i/L$ of these.

By similar reasoning, the total number of territories is given by
\begin{equation}
\Delta_+ N=(1-e^{-L})U=(1-e^{-\overline{b}N/T})(T-N), \label{eq:Nmaster}
\end{equation}
which is exact and applies for arbitrary $b_i$, $c_i$ and $d_i$.

@@

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be consistent between rare and common types. To illustrate, suppose that two identical types (same $b$, $c$ and $d$) are present, with low $l_1\ll 1$ and high density $l_2\approx L\gg 1$ respectively. Since $L$ is large, uncontested territories make up a negligible fraction of the total. The rare type's territorial acquisition is almost entirely due to $\Delta_r n_1$, while the common type's territorial acquisition entirely due to $\Delta_a n_2$. To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$. Even small violations of this consistency condition would mean exponential growth of one type relative to the other. This behavior is clearly pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes. Thus, predicted growth or decline would depend on an arbitrary assignment of rarity.

For example, suppose that we use $\tilde{p}$ and $\hat{p}$ to calculate the effective means. The right hand side of Eq. \eqref{eq:meanfieldr} is then approximately $1/(L+1)$, and since $l_1\ll 1$ and $L\gg 1$ we have $\Delta_r n_1 \approx 1/(L+1)$ in Eq.~\eqref{eq:deltr}. Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}} = L$ in Eq. \eqref{eq:meanfielda}, and so $\Delta_a n_2 \approx 1/L$. Thus, the identical rare type is  pathologically predicted to decline in frequency.

>>>>

In the special case that all types are competitively equivalent (identical $c_i$), Eq.~\eqref{eq:master} takes a simpler form,
\begin{equation}
\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U=\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N), \label{eq:masterequalc}
\end{equation}
where we have used the fact that $L=\overline{b}N/T$ to make the dependence on $b$ and $N$ explicit ($\overline{b}$ is the population mean $b$). Eq.~\eqref{eq:masterequalc} happens to be exact even though it is a special case of the approximation Eq.~\eqref{eq:master}. This can be deduced directly from Eq.~\eqref{eq:growthsumuncoupled}: $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of such territories, and type $i$ is expected to receive a fraction $l_i/L$ of these. By similar reasoning, the total number of territories acquired is given by
\begin{equation}
\Delta_+ N=(1-e^{-L})U=(1-e^{-\overline{b}N/T})(T-N). \label{eq:Nmaster}
\end{equation}
This formula is also exact, but unlike Eq.~\eqref{eq:masterequalc}, it also applies when the $c_i$ differ between types.

@@

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be consistent between rare and common types. To illustrate, suppose that two identical types (same $b$, $c$ and $d$) are present, that the first type is rare and has vanishingly small density $l_1\rightarrow 0$ while the second is common and has high density $l_2\rightarrow \infty$. Since $l_2\rightarrow \infty$, uncontested territories make up a negligible fraction of the total. The rare type's territorial acquisition is entirely due to $\Delta_r n_1$, while the common type's territorial acquisition is entirely due to $\Delta_a n_2$. To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must therefore be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$. Even small violations of this consistency condition would mean exponential growth of one type relative to the other. This behavior is clearly pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes. Thus, predicted growth or decline would depend on an arbitrary assignment of rarity.

In general, the above consistency condition is not satisfied when the conditional Poisson distributions $\tilde{p}$ and $\hat{p}$ are used to calculate the effective means. Under the above limits ($l_1\rightarrow 0$, $l_2\rightarrow \infty$), the right hand side of Eq.~\eqref{eq:meanfieldr} tends to $1/(l_2+1)$, and so $\Delta_r n_1/m_1 \rightarrow 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}. Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}} = l_2$ in Eq. \eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \rightarrow 1/L$. Thus, the identical rare type is pathologically predicted to decline in frequency.
